\bf Title

### 2023-10-30 article/paper_padic_spec.tex
variables

Throughout the paper, $K$ refers to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \ZZ \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer.
For $k \in \N$, we write $O(\pi^k)$ for $\pi^k \OK.$
A typical example of $K$ as above is the field of $p$-adic numbers 
$\QQ_p$ (equipped with the $p$-adic valuation). For this example, we 
have $\OK = \ZZ_p$.
Another example is the field of Laurent series
$K=\QQ(\!(T)\!)$ equiped with $T$-adic valuation.
In this case,  $\OK = \QQ \llbracket T \rrbracket$.
We refer to \cite{Serre:1979} for a
general introduction to such fields
and to X.~Caruso's course on $p$-adic computations \cite{caruso_computations_2017}
for an introduction to effective computations over $p$-adic numbers.

@@

Let $\sym_d(\R)$ be the vector space of $d \times d$ real symmetric matrices. A matrix $M \in \sym_d(\R)$
is called {\it positive semidefinite} ($M \succeq 0$) whenever the quadratic
form $x \mapsto x^TMx$ is nonnegative for all $x\in \R^n$. By the Spectral Theorem, $M \succeq 0$
if and only if the eigenvalues of $M$ are nonnegative, and this is equivalent to all the principal minors
of $M$ being nonnegative. The set of positive semidefinite matrices, denoted $\sym_d^+(\R) \subset \sym_d(\R)$,
is a closed convex cone with non-empty interior in $\sym_m(\R)$.

@@

\begin{example}\label{ell3}
  Let $d=n=3$. The $3$-elliptope is the three-dimensional spectrahedron given by
  $$
  \mathcal{E}_3 :=
  \left\{
  x
  \in \R^3 :
  \begin{bmatrix}
    1 & x_1 & x_2 \\
    x_1 & 1 & x_3 \\
    x_2 & x_3 & 1
  \end{bmatrix}
  \succeq 0
  \right\}
  =
  \left\{
  x
  \in \R^3 :
  \begin{array}{r}
    3-x_1^2-x_2^2-x_3^2 \geq 0 \\
    1+2x_1x_2x_3-x_1^2-x_2^2-x_3^2 \geq 0
  \end{array}
  \right\}.
  $$
  The polynomials on the right are the (non-constant) coefficients of the univariate polynomial
  $p(t) = \det(A(x)+t I)$,
  whose roots are the opposites of the eigenvalues of the defining matrix. Equivalently $\mathcal{E}_3$
  is defined by positivity of the principal minors of the defining matrix:
  $$
  1-x_1^2 \geq 0, \,\,\, 1-x_2^2 \geq 0, \,\,\, 1-x_3^2 \geq 0 \,\,\, \text{ and } \,\,\, 1+2x_1x_2x_3-x_1^2-x_2^2-x_3^2 \geq 0
  $$
  \exend
\end{example}

The optimization problem of miminizing a linear function over a real spectrahedron is called {\it semidefinite programming} (SDP). Given $C,A_0,A_1,\ldots,A_n \in \sym_m(\R)$, the corresponding semidefinite program is defined as
\begin{equation}
  \label{SDP}
\begin{array}{rcll}
  p^* & := & \inf_{x \in \R^n} & \left\langle c, x \right\rangle \\
  &    & \text{s.t.}         & A_0+\sum_{i=1}^n x_i A_i \succeq 0.
\end{array}
\end{equation}
Real polyhedra (subsets of $\R^n$ defined by finitely many affine inequalities) are special cases of real
spectrahedra, where the defining matrices $A_0, \ldots, A_n$ can be chosen diagonal. Thus SDP is a generalization
of linear programming (LP). In a fixed-precision model of computation, SDP is solvable in polynomial time in the
input size (matrix size,
number of variables, bit size of the coefficients), in $\log(\frac{1}{\varepsilon})$ (where $\varepsilon$ is the
precision) and $\log(\frac{1}{R})$, where $R$ is an {\it a priori} upper bound on the norm of the solution,
cf. \cite[Sec.1.9]{deKlerk}. In exact arithmetic, the complexity status of SDP is essentially open but complexity
bounds are known based either on general quantifier elimination \cite{ramana1997exact,porkolab1997complexity}
or by exploiting the determinantal structure of the constraints \cite{henrion2016exact}.

A more general class of convex semialgebraic sets are obtained by linear projections of real spectrahedra
called {\it semidefinite representable sets} (or spectrahedral shadows): these are sets of the form
$$
P = \left\{x = \left[\begin{smallmatrix} x_1 \\ \vdots \\ x_n \end{smallmatrix}\right] \in \R^n : \exists\,y\in\R^m, \, A_0 + \sum_{i=1}^n x_i A_i + \sum_{j=1}^m y_j B_j \succeq 0\right\}
$$

\begin{example}
\label{fermat_quartic}
The basic closed semialgebraic set $P = \{(x_1,x_2) \in \R^2 : 1-x_1^4-x_2^4 \geq 0\}$ is not a spectrahedron
but it is semidefinite representable:
$$
P = \left\{x = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix} \in \R^2 :
\exists\, (y_1,y_2) \in \R^2, \,
\begin{bmatrix}
  1+y_1 & y_2 \\
  y_2 & 1-y_1
\end{bmatrix}
\bigoplus
\begin{bmatrix}
  1 & x_1 \\
  x_1 & y_1
\end{bmatrix}
\bigoplus
\begin{bmatrix}
  1 & x_2 \\
  x_2 & y_2
\end{bmatrix}
\succeq 0
\right\}
$$
\end{example}

@@

In this section, we define polyhedra over a complete discrete valuation field $K$ with valuation $\val$,
and we discuss the associated linear optimization problem.

\begin{definition}
  Let $\ell_1,\ldots,\ell_d \in K[X]_1$ be linear inequalities on $X = (X_1, \ldots, X_n)$.
  A {\it polyhedron} is a subset $P \subset K^s$ defined by linear inequalities:
  $P = \val\left( \ell_1(x)\right)\ge 0$,\ldots, $ \val\left( \ell_n\left( x \right) \right)\ge 0$ for $ \ell_1,\ldots, \ell_n $ fixed linear forms and for all $x \in K ^s$.
\end{definition}

\begin{example}
	The closed unit ball of $K^s$ for the infinity norm [not sure if it is really used ] is convex polyhedron.It is the set of $\left( x_1,\ldots,x_s \right) \in  ^s$ so that $$
	\begin{pmatrix} x_1 &  & & \\
		& x_2 &0 &  \\
		&  0& \ddots & \\
		&  & & x_s/
	\end{pmatrix} \ge 0.$$ which is in fact $\OK^s$.

\end{example}

%\subsection{linear programming}
	We call { \it $p$-adic linear programming} the problem :
\begin{equation}
	\tag{PLp}
	\begin{matrix}
		\text{Minimize } \val\left(\left<c,x \right>\right) \text{ subject to }\\
		Ax + b \ge 0
	\end{matrix}
	\label{eqn:Proglinp}
\end{equation}

with $x$ a vector of $K ^n$, and $c$ a fixed vector of size $n$ that represents the cost, $A$ a fixed $m \times n $ matrix and $b$ fixed a vector of size $m$.

The method used in this article consists in reducing the Smith normal form of the matrix $A$.

\begin{definition} Smith normal form

	We call {\it Smith normal form} of a matrix $M \in \mathcal{M}_{m,n}\left(K \right) $ of rank $r$ the only matrix $S$ of the form $$S =  
	\begin{pmatrix} p^{a_1} & \\
		& \ddots \\
		& & p^{a_r}\\
		& & & 0\\
		& & & & \ddots \end{pmatrix} $$
	so that $(a_1)_{1\le i\le r}$ is a increasing sequence of integers and $M$ can be written $M =  Q^{-1} S P$ for $P \in \mathcal{G}L_n\left( \ZZ_p \right) $, $Q \in \mathcal{G}L_m\left( \ZZ_p \right) $.
\end{definition}

>>>>

Throughout the paper, $K$ refers to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \ZZ \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer.
For $k \in \N$, we write $O(\pi^k)$ for $\pi^k \OK.$
A typical example of $K$ as above is the field of $p$-adic numbers 
$\QQ_p$ (equipped with the $p$-adic valuation). For this example, we 
have $\OK = \ZZ_p$.
Another example is the field of Laurent series
$K=\QQ(\!(t)\!)$ equiped with $t$-adic valuation.
In this case, $\OK = \QQ \llbracket t \rrbracket$.
We refer to \cite{Serre:1979} for a
general introduction to such fields
and to X.~Caruso's course on $p$-adic computations \cite{caruso_computations_2017}
for an introduction to effective computations over $p$-adic numbers.

The $K$-vector space of polynomials of degree $\leq d$ on $x=(x_1,\ldots,x_n)$
with coefficients in $K$ is denoted by $K[x]_{d}$. The vector space of matrices of
size $p \times q$ with entries in $K$ is denoted by $\allmat_{p,q}(K)$. For the sake
of ease of writing, block-diagonal matrices with blocks $B_1,\ldots,B_d$ are denoted by
$B_1 \oplus \cdots \oplus B_d$.

@@

Let $\sym_d(\R) \subset \allmat_{d,d}(\R)$ be the vector space of $d \times d$ real symmetric
matrices. A matrix $M \in \sym_d(\R)$
is called {\it positive semidefinite} ($M \succeq 0$) whenever the quadratic
form $x \mapsto x^TMx$ is nonnegative for all $x\in \R^n$. By the Spectral Theorem, $M \succeq 0$
if and only if the eigenvalues of $M$ are nonnegative, and this is equivalent to all the principal minors
of $M$ being nonnegative. The set of positive semidefinite matrices, denoted $\sym_d^+(\R) \subset \sym_d(\R)$,
is a closed convex cone with non-empty interior in $\sym_d(\R)$.

@@

\begin{example}\label{ell3}
  Let $d=n=3$. The $3$-elliptope is the three-dimensional spectrahedron given by
  $$
  \mathcal{E}_3 :=
  \left\{
  x
  \in \R^3 :
  \begin{bmatrix}
    1 & x_1 & x_2 \\
    x_1 & 1 & x_3 \\
    x_2 & x_3 & 1
  \end{bmatrix}
  \succeq 0
  \right\}
  =
  \left\{
  x
  \in \R^3 :
  \begin{array}{r}
    3-x_1^2-x_2^2-x_3^2 \geq 0 \\
    1+2x_1x_2x_3-x_1^2-x_2^2-x_3^2 \geq 0
  \end{array}
  \right\}.
  $$
  The polynomials on the right are the (non-constant) coefficients of the univariate polynomial
  $p(t) = \det(A(x)+t I_3)$,
  whose roots are the opposites of the eigenvalues of the defining matrix. Equivalently $\mathcal{E}_3$
  is defined by positivity of the principal minors of the defining matrix:
  $$
  1-x_1^2 \geq 0, \,\,\, 1-x_2^2 \geq 0, \,\,\, 1-x_3^2 \geq 0 \,\,\, \text{ and } \,\,\, 1+2x_1x_2x_3-x_1^2-x_2^2-x_3^2 \geq 0
  $$
  \exend
\end{example}

The optimization problem of miminizing a linear function over a real spectrahedron is called {\it semidefinite programming} (SDP). Given $c \in \R^n$ and $A_0,A_1,\ldots,A_n \in \sym_d(\R)$, the corresponding semidefinite program is defined (in standard dual form) as
\begin{equation}
  \label{SDP}
\begin{array}{rcll}
  p^* & := & \inf_{x \in \R^n} & \left\langle c, x \right\rangle \\
  &    & \text{s.t.}         & A_0+\sum_{i=1}^n x_i A_i \succeq 0.
\end{array}
\end{equation}
Real polyhedra (subsets of $\R^n$ defined by finitely many affine inequalities) are special cases of real
spectrahedra, where the defining matrices $A_0, \ldots, A_n$ can be chosen diagonal, so that $A(x) =
\ell_1(x) \oplus \cdots \oplus \ell_d(x)$. In other words, SDP is a generalization of Linear Programming (LP).

In a fixed-precision model of computation, SDP is solvable in polynomial time in the input size (matrix size,
number of variables, bit size of the coefficients), in $\log(\frac{1}{\varepsilon})$ (where $\varepsilon$ is the
precision) and $\log(\frac{1}{R})$, where $R$ is an {\it a priori} upper bound on the norm of the solution,
cf. \cite[Sec.1.9]{deKlerk}. In exact arithmetic, the complexity status of SDP is essentially open but complexity
bounds are known based either on general quantifier elimination \cite{ramana1997exact,porkolab1997complexity}
or by exploiting the determinantal structure of the constraints \cite{henrion2016exact}.

A more general class of convex semialgebraic sets are obtained by linear projections of real spectrahedra
called {\it semidefinite representable sets} (or spectrahedral shadows): these are sets of the form
$$
P = \left\{x = \left[\begin{smallmatrix} x_1 \\ \vdots \\ x_n \end{smallmatrix}\right] \in \R^n : \exists\,y\in\R^p, \, A_0 + \sum_{i=1}^n x_i A_i + \sum_{j=1}^p y_j B_j \succeq 0\right\}
$$

\begin{example}
\label{fermat_quartic}
The basic closed semialgebraic set $P = \{(x_1,x_2) \in \R^2 : 1-x_1^4-x_2^4 \geq 0\}$ is not a spectrahedron
but it is semidefinite representable:
$$
P = \left\{\begin{bmatrix} x_1 \\ x_2 \end{bmatrix} \in \R^2 :
\exists\,
\begin{bmatrix} y_1 \\ y_2 \end{bmatrix} \in \R^2, \,
\begin{bmatrix}
  1+y_1 & y_2 \\
  y_2 & 1-y_1
\end{bmatrix}
\bigoplus
\begin{bmatrix}
  1 & x_1 \\
  x_1 & y_1
\end{bmatrix}
\bigoplus
\begin{bmatrix}
  1 & x_2 \\
  x_2 & y_2
\end{bmatrix}
\succeq 0
\right\}
$$
\end{example}

@@

In this section, we define polyhedra over a (complete) discrete valuation field $K$ with valuation $\val$,
and we discuss the associated linear optimization problem.

\begin{definition}
  A {\it polyhedron} is a subset $P \subset K^n$ of the form
  $P = \{x \in K^n : \val(\ell_1(x))\geq0,\ldots,\val(\ell_n(x))\geq 0\}$,
  for some $\ell_1,\ldots,\ell_d \in K[x]_1$. This will often be denoted with
  the compact notation $\val(\ell_1 \oplus \cdots \oplus \ell_d) \geq 0$, where $\val$
  and $\geq$ are applied entrywise.
\end{definition}

\begin{example}
  The closed unit ball $B_\infty(K^n)$ in $K^n$ for the supremum norm $||\cdot||_{\infty}$ is the
  polyhedron defined as the set of $x=(x_1,\ldots,x_n) \in K^n$ such that $||x||_{\infty} =
  \max_i\{x_i\} \leq 1$, in other words, such that $\val(x_i) \geq 0$, for all $i=1, \ldots, n$.
  In other words, $\OK^n = B_\infty(K^n)$ is the polyhedron in $K^n$ defined by the affine polynomials
  $\ell_i(x)=x_i$, $i=1,\ldots, n$.
%  $$
%  \begin{pmatrix} x_1 &  & & \\
%    & x_2 &0 &  \\
%    &  0& \ddots & \\
%    &  & & x_s/
  %  \end{pmatrix} \ge 0.$$
\end{example}

%\subsection{linear programming}

A {\it linear programming problem} (or simply {\it linear program}) is given by the following
optimization problem:
\begin{equation}
  \tag{LP}\label{LP}
\begin{array}{rcll}
  p^* & := & \inf_{x \in \R^n} & \val(\left\langle c, x \right\rangle) \\
  &    & \text{s.t.}         & A x + b \geq 0
\end{array}
\end{equation}
with $c \in K^n$ a fixed vector of size $n$ that represents the cost, $A \in \allmat_{m,n}(K)$, $b \in K^m$,
and $\left\langle c, x \right\rangle := c_1x_1+\cdots +c_nx_n$.
We present an algorithm for solving \eqref{LP} which consists in reducing the matrix $A$ in Smith normal form.
We recall its definition below, in our context.

\begin{definition}[Smith normal form] \label{smith_nf}
  Let $M \in \allmat_{m,n}(K)$ be of rank $r$. There exist unique integers $a_1 \leq \cdots \leq a_r$, and
  unique matrices $P \in GL_n(\OK)$ and $Q \in GL_m(\OK)$, such that
  $$
  S := QMP^{-1} = p^{a_1} \oplus p^{a_2} \oplus \cdots \oplus p^{a_r} \oplus {\bf 0}_{m-r}
  $$
  where ${\bf 0}_{m-r}$ is a null square matrix of size $m-r$.
  The matrix $S$ is called the {\it Smith normal form} of $M$.
\end{definition}
